Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-04-30 paper/ms.tex
ditto: propagate to ms

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission in order to derive rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages for our sample of FGK stars.
  %
  While the independent lithium and rotation-based ages agree
  for $\approx$90\% of stars, the timescales of lithium depletion are
  such that rotation is the more precise age diagnostic for the
  majority of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  For both planet-hosts, and for the broader sample of stars in the
  Kepler field, we find that the stellar age distribution decreases
  linearly from 3\,Gyr ago to present day, implying a linearly
  decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

To select stars with rotation periods, we turn to previous work.
Studies of stellar rotation across the entire Kepler field include
those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
and \citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach based on the
autocorrelation function to detect 34{,}030 rotation periods for
main-sequence Kepler targets cooler than 6500\,K, and excluded known
eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
candidates, or pulsators; they reported primary rotation periods for
24{,}124 of these stars.  \citet{Reinhold_2015} also reported
secondary periods for two thirds of the stars with primary periods,
which could be caused by differential rotation or finite spot
lifetimes.  Finally, \citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  The
rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} report rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes, and longer periods, than
those reported by \citetalias{McQuillan_2014}.  A relevant flag for
stars missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.  Reviewing the resulting
period detections, we learned that 32 KOIs with orbital and rotation
periods within $\approx$20\% were excluded from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, priv.~comm.); we choose to include these
objects.\footnote{In general, excluding stars in this category would
impose a bias against young close-in planets, due to the natural
commensurability between the rotation rates of young stars, and
Kepler's increased sensitivity to short-period objects.  While some of
these KOIs are known false positives, our analysis naturally excludes
such objects at later steps.  As one motivating example, this subset
of stars happens to include {\it e.g.}~Kepler~1643 (KIC~8653134), with
$P_{\rm orb}$=5.34\,days and $P_{\rm rot}$=5.05\,days, which is known
to be $\approx$40\,Myr old based on membership in RSG-5
\citep{Bouma_2022b}.  } The final tally of stars from
\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.

%  %TODO is it?
%  We verified that the main conclusions of our analysis do not change if
%  we were to adopt only the periods from \citetalias{McQuillan_2014}.
%  %TODO have you?

>>>>

% <250 words
\begin{abstract}
  Recent studies of the rotation rates and lithium content of FGK
  stars in open clusters have revealed how these quantities evolve
  during the first few billion years of a star's life.
  %
  Here we apply this knowledge to stars observed by Kepler in order to
  derive rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages for our sample of FGK stars.
  %
  The independent lithium and rotation-based ages agree for
  $\approx$90\% of stars.  However, the timescales of lithium
  depletion are such that rotation is the more precise age diagnostic
  for the majority of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  More broadly, we find that the age distribution of stars in the
  Kepler field decreases linearly from 3\,Gyr ago to present day,
  for both known planet hosts as well as the parent stellar sample.
  %
  While this stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

To select stars with rotation periods, we turn to previous work.
Many investigators have derived rotation periods both specifically of
the Kepler Objects of Interest
\citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
and also of the broader set of all Kepler target stars
\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
These studies used a range of detection methods and selection
functions.
We are interested in understanding the age distribution of not only
the KOIs but also the Kepler target stars.
Out of these studies, the only homogeneous analysis of both Kepler
targets and KOIs were the studies by
\citet{Santos_2019} (\citetalias{Santos_2019}),
and \citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
They included known KOIs and binaries in their analysis, and assigned
them specific quality flags, while removing transits and eclipses
during the stellar rotation measurement process.

%Studies of stellar rotation across the entire Kepler field include
%those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
%\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
%and \citet{Santos_2021} (\citetalias{Santos_2021}).
%\citetalias{McQuillan_2014} used an approach based on the
%autocorrelation function to detect 34{,}030 rotation periods for
%main-sequence Kepler targets cooler than 6500\,K, and excluded known
%eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
%Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
%variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
%candidates, or pulsators; they reported primary rotation periods for
%24{,}124 of these stars.  \citet{Reinhold_2015} also reported
%secondary periods for two thirds of the stars with primary periods,
%which could be caused by differential rotation or finite spot
%lifetimes.  Finally, \citetalias{Santos_2019} and
%\citetalias{Santos_2021} combined a wavelet analysis and
%autocorrelation-based approach, and cumulatively reported rotation
%periods for 55{,}232 main-sequence and subgiant FGKM stars.
%\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
%KOIs and binaries, and assigned them specific quality flags.  The
%rotation periods of KOIs have received considerable additional
%scrutiny
%\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}.  A relevant flag for
stars missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.
This is not surprising, in that the rotation periods of KOIs have
received considerably more scrutiny than those of ordinary Kepler
stars.
We therefore decided to split our subsequent analysis into a
homogeneous portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.
For the latter portion, we first included 32 KOIs with orbital and
rotation periods within $\approx$20\% that had been
excluded\footnote{Excluding stars in this category would impose a bias
against young close-in planets, due to the natural commensurability
between the rotation rates of young stars, and Kepler's increased
sensitivity to short-period objects.  Although some of these KOIs are
known false positives, our analysis excludes such objects at later
steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, private communication).
We then incorporated an additional \nnewdavidtwentyone\ rotation
periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.
Inclusion of these additional KOI rotation periods is a
supplementary measure aimed at completeness in our final KOI age
catalog; the provenances of the individual adopted periods are all
noted in the relevant tables.

Finally, since our scope is focused on rotation-based age dating, 
we restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.
FGK stars with slower rotation periods spin much slower than the stars
in the oldest open clusters currently used to calibrate empirical
gyrochronal models.

The resulting tally of stars from
\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.
